Hydraulically-vulnerable trees are buffered by deep-water access in a tropical forest community

### 2020-10-26 old_ms_sections.Rmd
moved in lots of text to fit in word count limit

@@

From Discussion first para

@@

>>>>

To estimate effective rooting depth, we tested alternate empirical models for species 5-yrly census growth (Eq. \@ref(eq:M-chosen) and those in SI), which use effective rooting depth as an input. Many intrinsic and extrinsic factors limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019]. Rhizosphere soil water potential ($\mathrm{\Psi}$~soil~) down-regulates $\mathrm{\Psi}$~leaf~ [@Christoffersen:2016kd2d] such that in moist forests $\mathrm{\Psi}$~leaf~ equilibrates with $\mathrm{\Psi}$~soil~ at night, declining through the day. Using species-specific leaf hydraulic vulnerability curves, $\mathrm{\Psi}$~leaf~ can be converted to $K_\textrm{leaf}$, which in turn controls growth [@Sack.2012]. We therefore used maximum diurnal ($K_\textrm{leaf}$) determined from $\mathrm{\Psi}$~soil~ as a proxy for the soil water limitation on growth. $K_\textrm{leaf}$ in our models thus incorporates the effects of rooting depth and rhizosphere water potential. This approach, however, does not account for the buffering effect of stem water storage capacitance on $\mathrm{\Psi}$~leaf~, which may cause the model to slightly overestimate ERD (see Discussion). VPD affects growth non-linearly, with growth increasing with VPD upto a VPD threshold, then declining as leaf pores (stomata) close [@Grossiord.2020]. Apart from stomatal control, leaf-drop may further limit water-uptake and growth. We therefore regressed species-specific growth against multiplicative or additive effects of one or more factors calculated daily and averaged over 5-yrly censuses: VPD, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1. Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting. At our study site, due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season, the impact of VPD may as well be a combined impact of not only temperature and humidity, but also radiation.

@@

#Discussion

@@

Our result of deep rooted species being more hydraulically vulnerable is consistent with observations for a seasonal forest in the Amazon . @Johnson:2018fad found that mortality decreased with deep rooting even though deep rooted species were more vulnerable to xylem cavitation (high $\mathrm{\Psi}$~88,stem~) and had narrower safety-margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~). @Nardini:2015 found that a species sourcing deep water in a Karst system showed lower hydraulic failure and mortality in an extreme drought despite having high $\mathrm{\Psi}$~88,stem~ and narrower safety margin compared to two species sourcing shallow-water.

at Mudumalai is nearly one (Penman-Monteith PET = 980 $\pm$ 64 mm)
and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (Penman-Monteith PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (Penman-Monteith PET = `r round(mean(clim$pet.PM))` $\pm$ `r round(sd(clim$pet.PM))` mm;`r min(clim$Year)`-`r max(clim$Year)`; this study, @Guo.2016),

\
\   We found a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA and WSG) and parameters of leaf vulnerability (or percent loss of conductance, PLC) curves, a 'hard' trait, based on one of the largest dataset for the tropics ($n$ = `r length(sp.exp.param$sp)`). This allowed us to scale-up the leaf vulnerability curves at the community scale. We find that leaf conductance declined more rapidly with declining $\mathrm{\Psi}$~leaf~ for species with higher values of WSG and LMA (Fig. S\@ref(fig:obs-plc-curves)). Whether such scaling relationships between hard and soft traits hold true for other forests, and their wider applicability, remains to be seen as data on PLC curves are limited in tropical forests, but supportive and counter examples both exist [@McCulloh.2011; @Hoffmann.2011]. WSG alone explained `r round(b.wsg.r2*100,0)`% of the variance in $B$ ($p$ `r  b.wsg.p`).

However, trees' stem diameter change is a sum total of not just growth, the addition of wood, but also changes due to reversible stem dehydration (xylem, phloem and bark), the importance of which increases with temporal resolution [@Chitra-Tarak.2016; @Chitra-Tarak.2015].

Use of dendrometer band data would thus entail accounting for reversible stem dehydration

This would also open the possibility of testing, whether deciduous species are more shallow rooted than evergreen species as generally observed [@Fan:2017]

, and suggests that greater emphasis should be given to predictive modeling of tree root allocation and distribution (e.g. @Wijk.2001) as well as trees' hydrological environments

\
\ Finally, rooting depths may be parametarized within DGVMs such as ELM-FATES using plant hydrodynamic schemes [@Christoffersen:2016kd2d; @Koven.2020], simultaneously modeling plant water-availability with plant water-extraction. In general, including differences in tree rooting depths in ELM-FATES would improve understanding of plant competition for water (and other resources) and more accurately represent plant's drought strategies, leading to better predictions of existing and future forest function and composition.
